---
title: "Replicating the Ecological Impact of High-Speed Rail: A Staggered Difference‑in‑Differences Tutorial"
author: "Ixel Medrano, Hylaea Miller, Melannie Moreno Rolon"
format: html
date: "March 06, 2026"
---

### Setup: Load libraries

0.  Load libraries (+ install if needed)

```{r}
library(tidyverse)
library(here)
library(janitor)
library(jtools)

library(gtsummary)
library(gt)

library(MatchIt) # matching
library(cobalt)  # balance + love plots

library(fixest) # fast fixed effects
library(scales) # plotting
```

### Part 1: Study Background


```{r}
match_data <- read_csv(here::here("data", "xie2025_simulated_panel.csv" )) %>% mutate(region=factor(region), city_scale=factor(city_scale), lat_band=factor(lat_band))
```
#### **2B.** Check imbalance (before matching)

-   Create a covariate *balance table* comparing treated and control precincts
-   Treatment indicator: `City-year = 1` if HSR is operating in that city that year
0 otherwise
`
-   Include pre-treatment covariates: `log_home_val_07`, `p_uni_degree`, `log_median_inc`, `log_pop_denc`
-   Use the `tbl_summary()` function from the `{gtsummary}` package.


```{r}
match_data %>%
  select(
    hsr,
    gdp_growth_pct,
    pm25,
    temperature_c,
    population_10k
  ) %>%
  tbl_summary(
    by = hsr,   # treatment
    statistic = list(
        all_continuous() ~ "{mean} ({sd})",
        all_categorical() ~ "{n} ({p}%)"
  )) %>%
  modify_header(label = "Covariate") %>% 
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Group**")
```

```{r}
set.seed(2412026)

match_model <- matchit(
  hsr ~ gdp_growth_pct + pm25 + temperature_c + population_10k,
  data = match_data %>% 
    drop_na( gdp_growth_pct, pm25, temperature_c, population_10k),
  method = "nearest",       # Nearest neighbor matching
  distance = "mahalanobis", # Mahalanobis distance
  ratio = 1,                # Match one control unit to one treatment unit (1:1 matching)
  replace = FALSE           # Control observations are not replaced
)

# Extract matched data
matched_data <- match.data(match_model)

```
```{r}
summary(match_model)
```

```{r}
new_names <- data.frame(
    old = c("log_home_val_07", "p_uni_degree", "log_median_inc", "log_pop_denc"),
    new = c("Home Value (log)", "Percent University Degree",
            "Median Income (log)", "Population Density (log)"))

# Love plot
love.plot(match_model, stats = "mean.diffs",
          thresholds = c(m = 0.1),
          var.names = new_names)
```

