---
title: "Parallel test"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(fixest)
library(MatchIt)
library(modelsummary)
library(broom)
library(knitr)
library(here)

# Load library 

hsr_data <- read_csv(here("data", "xie2025_simulated_panel.csv")) %>%
  mutate(region = as.factor(region), 
         lat_band = as.factor(lat_band),
         city_scale = as.factor(city_scale), 
         hsr_open_year = na_if(
           hsr_open_year, Inf)) # Treat Inf as missing value
```

## Event Study & Parallel Trends

Note: A DiD is only valid if the Treatment and Control groups were moving in the same direction before the intervention hit. Let's see if we pass the parallel trends assumption.

The parallel trends assumption is untestable in a strict sense — we can never observe what treated cities would have looked like without HSR. But we can check whether treated and control cities were on **similar trajectories before** treatment. If pre-treatment coefficients are close to zero and statistically insignificant, this supports the assumption.

The event study model replaces the single `hsr` indicator with a set of dummies for each year relative to HSR opening:

$$\text{Avian}_{it} = \alpha_0 + \sum_{b \neq -3} \beta_b \cdot \mathbf{1}[\text{event\_time}_{it} = b] \times \text{Treated}_i + \gamma X_{it} + \mu_i + \lambda_t + \epsilon_{it}$$

Each $\beta_b$ captures the gap between treated and control cities $b$ years relative to HSR opening, compared to the reference year (here, 6 years before opening).

```{r event-study}
hsr_es <- hsr_data %>% 
  mutate(event_time_binned = case_when(
    is.na(event_time) ~ NA_real_,    # control cities stay as NA
    event_time < -6   ~ -6,          # bin anything 6+ years before
    event_time >  6   ~  6,          # bin anything 6+ years after
    TRUE              ~ event_time
  ))
```

 Run the event study regression**

Instead of a single `hsr` dummy, `i(event_time_binned, ref = -6)` automatically creates one dummy per relative year. The `ref = -6` argument drops year −6 from the regression, making it our baseline that all other years are compared against.

```{r event-study}
es_model <- feols(
  avian_increment_ratio_natl ~
    i(event_time_binned, ref = -6) +
    gdp_growth_pct + share_primary_pct + share_secondary_pct +
    population_10k + urban_road_10k_m2 + pm25 + green_coverage_pct +
    wind_speed + precip_mm + humidity_pct + temperature_c |
    city_id + year,
  data    = hsr_es,
  cluster = ~city_id
)

# Extract coefficients with tidy
es_coefs <- tidy(es_model, conf.int = TRUE) %>% 
  filter(str_detect(term, "event_time_binned")) %>%  # only want event-time coefficients 
  mutate(event_time = as.numeric(str_extract(term, "-?\\d+"))) # regex pattern to help extra negative value to use on x-axis  
%>% 
  bind_rows(tibble(event_time = -3, estimate = 0, conf.low = 0, conf.high = 0)) %>% 
  arrange(event_time)
```

# Plot the event study

Each dot is a coefficient estimate. Pre-treatment dots (left of the dotted line) should sit near zero — that is the parallel trends check. Post-treatment dots should turn negative, showing HSR's impact on bird species influx.

```{r event-study}
es_coefs %>% 
  ggplot(aes(x = event_time, y = estimate)) +
  annotate("rect", xmin = -5.5, xmax = -0.5,
           ymin = -Inf, ymax = Inf, alpha = 0.04, fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = -0.5, linetype = "dotted",
             color = "steelblue", linewidth = 0.7) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 2.5) +
  annotate("text", x = -3, y = max(es_coefs$conf.high) * 0.15,
           label = "ref", size = 3, color = "grey40") +
  annotate("text", x = 0.2, y = max(es_coefs$conf.high) * 0.85,
           label = "HSR opens →", hjust = 0, size = 3.5, color = "steelblue") +
  scale_x_continuous(breaks = -5:5) +
  labs(
    title    = "Event Study: Effect of HSR on Bird Species Increment",
    subtitle = "Pre-treatment coefficients near zero support the parallel trends assumption",
    x        = "Years Relative to HSR Opening",
    y        = "Avian Increment Ratio (National)",
    caption  = "Two-way FE (city + year), full controls, SEs clustered at city level."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold"),
    plot.subtitle    = element_text(color = "grey40", size = 10),
    panel.grid.minor = element_blank()
  )
```

### Reading the plot

- **Left of the dotted line (pre-treatment):** coefficients should hover near zero — this is the parallel trends check. Any significant pre-trend would suggest the two groups were already diverging before HSR, undermining the causal interpretation.
- **Right of the dotted line (post-treatment):** coefficients should become negative and significant, consistent with HSR reducing bird species influx.
- **The paper's key finding:** the decline begins *two years before* HSR opens, corresponding to the construction phase. This is an important substantive result — environmental damage starts before trains run.
